---
title: Multi-ancestry finemapping of gene clusters implicated in cardiovascular disease and mental health
author: Mark James Adams, INCR
date: 2026-01-31
execute:
  echo: false
  warning: false
  message: false
format:
  pdf:
    sansfont: "Crimson Pro"
    df-print: kable
reference-location: margin
citation-location: document
bibliography: ref.bib
csl: human-genetics.csl
---

<!-- layout information: https://quarto-dev.github.io/quarto-gallery/page-layout/tufte.html -->

A genomic region containing a cluster of genes that regulate fatty acid desaturation has recently been found to have a strong association with major depressive disorder [@Adams_MDD2025_Cell]. However, it is unclear which variant in this region is the cause of the association. This project will use biobank whole genome-sequencing datasets to finemap common variants, catalogue rare variants, and understand shared genetic risk between cardiovascular disease and common psychiatric disorders.

```{r library}
library(dplyr)
library(readr)
library(arrow)
library(topr)
```

```{r sumstats}
mdd2025 <- read_parquet(here::here("data/raw/sumstats/daner_pgc_mdd_full_div_hg19_v3.49.46.01.neff.parquet"))
```

# The FADS gene cluster

The fatty acid desaturase gene family (FADS1, FADS2, FADS3) is co-located in a gene-dense region on chromosome 11 (11q12-q13.1). FADS1 and FADS2 regulate the conversion of polyunsaturated fatty acids (omega-6 and -3) into long-chain forms, which have roles in a wide range of cellular processes and are implicated in cardiovascular, metabolic, and immunological health [@Koletzko_FADS_2019].

The role of FADS as the genes underlying the association with major depression is bolstered by a genetic causal analysis (Mendelian randomisation) of circulating metabolites [@davyson_metabolomic], which showed that omega-3 levels and omega-6 to omega-3 ratios had a causal effect on depression. While this makes FADS1 and FADS2 the most plausible biological candidates, expression levels of other genes in this region (MYRF and TMEM258) also have possible links to depression. Further work is therefore required to pinpoint the causal variant driving the association.

## Genetic associations

Prior studies examining specific FADS genetic variants and depression [@Cribb14092018;@Sublette_FADS_2016] had small sample sizes (N $\approx$ 200–1000). Large genome-wide association studies (GWASs) have identified a strong association signal in this region, but the most significant variant has been located in other nearby genes like DAGLA [@howard.d.adams.m.ea:genome-wide;@meng:gwas:2024] and TMEM258 [@levey:mvp:2021;@als:gwas:2023]. However, the most recent and largest multi-ancestry genetic study of depression found the top signal to be a single nucleotide intron variant in FADS2 (@fig-fads_mdd2025). 



```{r}
#| label: fig-fads_mdd2025
#| cache: true
#| fig-cap: "Region plot of genomewide association findings for the FADS gene cluster. Top: Tested variants with transformed association p-value. Bottom: Gene positions in the region."
#| fig-width: 11
#| fig-height: 6
#| fig-column: page-right
regionplot(
  mdd2025,
  region = "11:61380730-61678348",
  build = 37,
  show_overview = FALSE,
  rsids_with_vline = c("rs28456", "rs174544"),
  rsids_color = "red",
  unit_main = 7,
  unit_gene = 4,
  verbose = FALSE
)
```

## Finemapping

Because genetic variants in a region are in close physical linkage, association studies find a correlated statistical signal across variants that makes it difficult to determine which variant is the cause of the association. Finemapping is the process of comparing the linkage and association signals to determine the probability that each SNP is the causal SNP. Finemapping can be enhanced by considering prior causal probabilities based on functional annotations (such as a coding or regulatory function, or evidence of mutational constraint).

Previous investigations of this region used (1) association summary statistics from European-ancestry cohorts only and (2) summary statistics from all ancestry groups merged together.  Pre-computed information from UK Biobank was used and and it was assumed that the region only had a single causal variant. This latter assumption was stipulated because it does not require that the linkage disequilibrium estimates be population accurate. The assumption of a single causal variant was also reasonable given that the conditional-and-joint analysis (which tests whether each other variant is still significant after controlling for the top associated variant in the region) did not indicate the presance of multiple causal variants.

However, this analysis did not identify a variant that met the finemapping identification criterion, which was a posterior inclusion probability (PIP) $> 0.95$. The most significant variant in the multi-ancestry analysis was rs28456 [^rs28456] while the finemapping analysis selected rs174544 [^rs174544] as the more likely causal SNP.

[^rs28456]: [rs28456](https://www.ncbi.nlm.nih.gov/snp/rs28456): an intron variant in the [FADS2](https://www.ncbi.nlm.nih.gov/gene/9415) gene and the uncharacterised gene [LOC124902680](https://www.ncbi.nlm.nih.gov/gene/?term=LOC124902680);  PIP = 0.122.
[^rs174544]: [rs174544](https://www.ncbi.nlm.nih.gov/snp/rs174544): a variant in $3^\prime$ untranslated region of [FADS1](https://www.ncbi.nlm.nih.gov/gene/3992); PIP = 0.181.




## Rare variant burden 

An analysis to look for the effect of rare variants on major depression and on metabolite levels has also been conducted [@nisbit:fads:wcpg2025]. This study used short read whole genome sequencing (srWGS) datasets from All of  Us and UK Biobank. It considered metabolic measures that had been associated with major depression (Omega 3, Omega 3 percentage, Docosahexaenoic acid , Omega 6/Omega 3 ratio and Degree of Unsaturation.).  The authors found that rare variant burden was significantly associated with all tested metabolite measures but was not associated with major depression. This indicates that the GWAS association is likely to caused by a common variant than being a proxy for a rare variant.

# Analyses

In this project I aimed to discover more about causal variants in the FADS region using mult-ancestry and whole-genome sequencing datasets. Initial analysis improved on previous methods in three ways:

- Rather than using summary statistics from merged ancestries, use summary statistics from the single largest ancestry group with a matched reference panel to test for multiple causal variants.
- Use disaggregated summary statistics from multiple ancestries to improve finemapping.
- Use a large, multi-ancestry-matched WGS dataset as an improve reference panel for finemapping. 

All analyses were focused on the FADS gene region `chr11:61400000-61700000` using summary statistics from Adams [-@Adams_MDD2025_Cell] and Meng [-@meng:gwas:2024]. The ancestry clusters used were dervied from continental reference populations, namely African (AFR), Admixed American/Hispanic/Latino (AMR), East Asian (EAS), European (EUR), and South Asian (SAS). I conducted finemapping using PolyFun [@Weissbrod:PolyFun:2020].

Finemapping attempts to identify "credible causal sets", which are a group of variants that collectively capture an association signal in a region of the genome. The variants are ranked by the a posterior inclusion probability (PIP), which combines the prior functional probabilites with the data to determine a posterior causal probability for each variant. A PIP > 0.95 is typically taken as sufficient to identify the causal SNP, but values less than this are still useful for narrowing down the set of potentially causal variants. 

## Test for multiple causal variants

Testing for multiple causal variants requires an ancestry-matched reference panel. I started with the standard reference panel from the 1000 Genomes Project [@the_1000_genomes_project_consortium_global_2015]. I used prior probability information from PolyFun [@Weissbrod:PolyFun:2020] and then performed finemapping using FINEMAP [@Benner:FINEMAP:2016] on the European-ancestry cluster association summary statistics (N cases = 525,197; N controls = 3,362,335).

```{r}
#| label: polyfun_eur_ref
polyfun_eur_ref <- read_tsv(here::here("reports/tables/PolyFun/mdd2025_eur_all_hg19_polyfun_finemap_chr11_61400000-61700000.tsv"))

# get credible sets
polyfun_eur_ref_cs <- polyfun_eur_ref |>
  filter(CREDIBLE_SET != 0)
```

The finemapping analysis found three credible causal sets which contained 6, 2, and 175 variants respectively (@tbl-finemap-eur-ref). While no SNP had a PIP > 0.95, this analysis supports the conclusion that there are multiple causal variants in the FADS region. The third credible causal set (CCS) contained the SNP (rs174544) that was the top finemapped variant in the published analysis while also containing MYRF [^MYRF] intron variants. The first CCS contained SNPs that were primarily in FADS2 introns, but there was one variant (rs174567) that was an upstream transcript variant. The second CCS contained intron variants in DAGLA.[^DAGLA]

[^MYRF]: [MYRF](https://www.ncbi.nlm.nih.gov/datasets/gene/745/): myelin regulatory factor.
[^DAGLA]: [DAGLA](https://www.ncbi.nlm.nih.gov/datasets/gene/747/): diacylglycerol lipase alpha, involved in biosynthesis of endocannabinoid ligands.


```{r}
#| label: tbl-finemap-eur-ref
#| tbl-cap: Top five SNPs with highest posterior inclusion probability (PIP) in each credible causal set from finemapping of European-ancestry cluster with 1000 Genomes reference panel.

polyfun_eur_ref_cs |>
  group_by(CREDIBLE_SET) |>
  arrange(CREDIBLE_SET, desc(PIP)) |>
  slice_head(n = 5)|>
  arrange(CREDIBLE_SET, BP) |>
  select( 
    `Credible set` = CREDIBLE_SET,
    SNP,
    `Position (b37)` = BP,
    PIP
  )
```

## Multi-ancestry finemapping

## WGS reference dataset


## Acknowledgments 

This work was supported by a BHF REA4 Pump Priming Award (RE/24/130012).

## References

::: {#refs}
:::